Election 2024

### 2024-10-31 paper/paper.qmd
prepare for merge

```{python}
#| echo: false
import pandas as pd
import matplotlib.pyplot as plt
import jupyter
from tabulate import tabulate
```

@@

The voter support data used in this analysis comes from raw polling information sourced from the Project 538 online database [@FiveThirtyEight]. This dataset provides predictions for candidate support for the 2024 U.S. Presidential Election by aggregating various polls that capture public sentiment towards candidates Donald Trump and Kamala Harris. Each entry in the dataset corresponds to a specific question around voting preference in a poll conducted by different polling organisations, which measure the percentage of respondents expressing their support for each candidate in their respective state.

@@

```{python}
import matplotlib.pyplot as plt
import numpy as np

# Step 1: Define the Electoral Votes for Each State
states = ["California", "Connecticut", "Delaware", "Hawaii", "Illinois",
"Maine", "Maryland", "Massachusetts", "Minnesota", "New Jersey",
"New York", "Oregon", "Rhode Island", "Vermon", "Washington",
"District of Columbia", "New Mexico", "New Hampshire", "Colorado", "Virginia",
"Arizona",
"Georgia",
"Michigan",
"Nevada",
"North Carolina",
"Pennsylvania",
"Wisconsin",
"Alabama", "Alaska", "Arkansas", "Florida", "Idaho",
"Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana",
"Mississippi", "Missouri", "Montana", "Nebraska", "North Dakota",
"Ohio", "Oklahoma", "South Carolina", "South Dakota", "Tennessee",
"Texas", "Utah", "West Virginia", "Wyoming"]
print(len(states))
votes = [54, 7 , 2, 4, 19, 4, 10, 11, 10, 14, 28, 8, 4, 3, 12, 3, 5, 4, 10, 13,
         11,16,15,6,16,19,10,
         9,3,6,30,4,11,6,6,8,8,6,10,4,5,3,17,7,9,3,11,40,6,4,3]

print(len(votes))
# Step 2: Compute the Cumulative Electoral Votes
cumulative_votes = np.cumsum(votes)

# Compute Cumulative Votes in Reverse Order for Republicans
cumulative_votes_reversed = np.cumsum(votes[::-1])

# X-axis values (state index)
x = np.arange(len(cumulative_votes))

# Step 3: Plot the Step Function for Democrats (left to right)
plt.figure(figsize=(18, 8))
plt.step(x, cumulative_votes, where='post', color='navy', linewidth=2, marker='o')

# Reverse the x-axis for the Republican step function (right to left)
x_reversed = np.flip(x)
plt.step(x_reversed, cumulative_votes_reversed, where='post', color='red', linewidth=2, marker='o')

# Step 4: Customize the Plot
plt.xticks(range(len(states)), states, rotation=90, fontsize=8)
plt.yticks(np.arange(0, 550, 50))

# Fill the area under the curve with blue for Democrats (left to right)
threshold_index = np.argmax(cumulative_votes >= 227)  # First point where cumulative votes reach 270
plt.fill_between(x[:threshold_index + 1], cumulative_votes[:threshold_index + 1], 
                 step='post', color='blue', alpha=0.4)

# Fill the area under the curve with red for Republicans (right to left)
threshold_index = np.argmax(cumulative_votes >= 219)
plt.fill_between(x_reversed[:threshold_index + 5], cumulative_votes_reversed[:threshold_index + 5], step='post', color='red', alpha=0.4)

# Add a horizontal line at 270 votes (threshold to win)
plt.axhline(270, color='purple', linestyle='--', linewidth=2)
plt.text(len(states) / 2, 275, '270 votes to win', horizontalalignment='center', color='purple', fontsize=12)

# Add lines for current Democratic and Republican thresholds
plt.axhline(225, color='blue', linestyle='--', linewidth=1.5)
plt.text(210, 220, '225 votes', horizontalalignment='center', color='blue', fontsize=12)

# Add Labels and Title
plt.xlabel('Electoral College Markets (electoral votes)', fontsize=12)
plt.ylabel('Total Electoral Votes', fontsize=12)
plt.title('Step Function of Total Electoral Votes\nwith States Ordered by Vote Share', fontsize=15)

# Add Grid for Better Visibility
plt.grid(True, linestyle='--', alpha=0.5)

# Adjust Layout for Better Fit
plt.tight_layout()

# Step 5: Show the Plot
plt.show()

@@

# Load your data (replace with your actual data file path)
data = pd.read_parquet('../data/02-analysis_data/merged_swing_state_data.parquet')

# Ensure date column is in datetime format
data['end_date'] = pd.to_datetime(data['end_date'])

# Plot the histogram of poll quantity by end date
plt.figure(figsize=(12, 6))
plt.hist(data['end_date'], bins=20, edgecolor='black', alpha=0.7)

# Adding labels and title
plt.xlabel('End Date of Poll')
plt.ylabel('Number of Polls')
plt.title('Histogram of Poll Quantity by End Date')
plt.xticks(rotation=45)

# Show the plot
plt.tight_layout()
plt.show()

@@

# 4 Results {#sec-results}
The results of our analysis are as follows. We have Donald Trump predicted to win Arizona, Georgia and Michigan for a total of 42 of the 93 swing state votes. While Kamala Harris is projected to win Nevada, North Carolina, Pennsylvania and Wisconsin for the remaining 51 swing state votes. This results in a win for the democrats and the first female US president!

@@

# Appendix A: Idealized Methodology and Survey for Forecasting the US Presidential Election with Incentivized Betting

@@

>>>>

```{python}
#| echo: false
import pandas as pd
import matplotlib.pyplot as plt
import jupyter
from tabulate import tabulate
import numpy as np
```

@@

The voter support data used in this analysis comes from raw polling information sourced from the Project 538 online database [@FiveThirtyEight]. This dataset provides predictions for candidate support for the 2024 U.S. Presidential Election by aggregating various polls that capture public sentiment towards candidates Donald Trump and Kamala Harris. Each entry in the dataset corresponds to a specific question around voting preference in a poll conducted by different polling organisations, which measure the percentage of respondents expressing their support for each candidate in their respective state. the surveys utilized one or multiple of the following method for rsponse gathering: Online Ad, Online Panel, Live Phone, text-to-web, mail-to-web, email, Probability panel, IVR. The diversity of the methods of collection supplies for a colorful array of polls capturing all facets of voters.

@@

```{python}
# Define states and electoral votes, ensuring votes are listed in reverse for clarity in plot
states = ["California", "Connecticut", "Delaware", "Hawaii", "Illinois", "Maine", 
          "Maryland", "Massachusetts", "Minnesota", "New Jersey", "New York", "Oregon", 
          "Rhode Island", "Vermont", "Washington", "District of Columbia", "New Mexico", 
          "New Hampshire", "Colorado", "Virginia", "Arizona", "Georgia", "Michigan", 
          "Nevada", "North Carolina", "Pennsylvania", "Wisconsin", "Alabama", "Alaska", 
          "Arkansas", "Florida", "Idaho", "Indiana", "Iowa", "Kansas", "Kentucky", 
          "Louisiana", "Mississippi", "Missouri", "Montana", "Nebraska", "North Dakota", 
          "Ohio", "Oklahoma", "South Carolina", "South Dakota", "Tennessee", "Texas", 
          "Utah", "West Virginia", "Wyoming"]
votes = [54, 7, 3, 4, 19, 4, 10, 11, 10, 14, 28, 8, 4, 3, 12, 3, 5, 4, 10, 13, 
         11, 16, 15, 6, 16, 19, 10, 9, 3, 6, 30, 4, 11, 6, 6, 8, 8, 6, 10, 4, 5, 
         3, 17, 7, 9, 3, 11, 40, 6, 4, 3]

# Calculate cumulative votes
cumulative_votes = np.cumsum(votes)
cumulative_votes_reversed = np.cumsum(votes[::-1])
x = np.arange(len(states))

plt.figure(figsize=(18, 8))
plt.step(x, cumulative_votes, where='post', color='navy', linewidth=2, marker='o')
x_reversed = np.flip(x)
plt.step(x_reversed, cumulative_votes_reversed, where='post', color='red', linewidth=2, marker='o')

# Threshold for filling
dem_threshold = np.argmax(cumulative_votes >= 227)
plt.fill_between(x[:dem_threshold + 1], cumulative_votes[:dem_threshold + 1], step='post', color='blue', alpha=0.4)

# Adding threshold line and markers
plt.axhline(270, color='purple', linestyle='--', linewidth=2)
plt.text(len(states) / 2, 275, '270 votes to win', horizontalalignment='center', color='purple', fontsize=12)
plt.xlabel('Electoral College by State', fontsize=12)
plt.ylabel('Total Votes', fontsize=12)
plt.title('Cumulative Electoral Votes by State', fontsize=15)
plt.grid(True, linestyle='--', alpha=0.5)
plt.tight_layout()
plt.show()

@@

# Load data (use an accurate path or dataset)
data = pd.read_parquet('../data/02-analysis_data/merged_swing_state_data.parquet')
data['end_date'] = pd.to_datetime(data['end_date'])

plt.figure(figsize=(12, 6))
plt.hist(data['end_date'], bins=20, edgecolor='black', alpha=0.7)
plt.xlabel('Poll End Date')
plt.ylabel('Poll Count')
plt.title('Poll Distribution Over Time')
plt.xticks(rotation=45)
plt.tight_layout()
plt.show()

@@

 # 4 Results {#sec-results}
 ## Regression results

 ```{Python}
#| echo: false

import pandas as pd
from tabulate import tabulate

# Data setup
data = {
    'State': ['Arizona', 'Georgia', 'Michigan', 'Nevada', 'North Carolina', 'Pennsylvania', 'Wisconsin'],
    'Trump % (Mean)': [49.85, 49.39, 48.57, 47.9, 48.48, 47.68, 47.76],
    'Trump % (SD)': [1.24, 1.78, 2.14, 1.5, 1.41, 1.81, 2.0],
    'Harris % (Mean)': [47.04, 47.66, 47.72, 48.63, 48.58, 48.53, 48.92],
    'Harris % (SD)': [2.17, 2.52, 2.12, 1.19, 1.55, 1.29, 1.72],
    'Percentage Chance of Trump Winning': [86.85, 71.18, 53.09, 21.34, 32.31, 19.67, 15.52]
}

# Create DataFrame
df = pd.DataFrame(data)

# Display as a professional table
table_output = tabulate(
    df, headers='keys', tablefmt='grid', showindex=False,
    floatfmt=(".2f", ".2f", ".2f", ".2f", ".2f", ".2f")
)
print(table_output)

in figure xx we cans see th results from the model described.

In Arizona and Georgia, the model predicts the strongest likelihood of a Trump victory, with a mean polling percentage showing Trump over 2 percentage points ahead in Arizona. In Georgia, Trump's mean polling percentage is 49.39% against Harris's 47.66%.The data is thus resulting in an strong 86.85% and and 71.18% likelyhood in Trump winning in Arizona and Georgia respectively.
Trump is also predicted to win in Michigan, alhough by smaller margins as shown in figure xx. As a rssult the model predicts trump winning with only a 53% chance. With Arizona, Georgia and Michigan Trump is predicted a total of 42 of the 93 swing state votes.

# Data setup (can adjust based on model results)
data = {
    'State': ['Arizona', 'Georgia', 'Michigan', 'Nevada', 'North Carolina', 'Pennsylvania', 'Wisconsin'],
    'Trump % (Mean)': [49.85, 49.39, 48.57, 47.9, 48.48, 47.68, 47.76],
    'Trump % (SD)': [1.24, 1.78, 2.14, 1.5, 1.41, 1.81, 2.0],
    'Harris % (Mean)': [47.04, 47.66, 47.72, 48.63, 48.58, 48.53, 48.92],
    'Harris % (SD)': [2.17, 2.52, 2.12, 1.19, 1.55, 1.29, 1.72],
    'Trump Win Probability (%)': [86.85, 71.18, 53.09, 21.34, 32.31, 19.67, 15.52]
}
df = pd.DataFrame(data)
df['Distance from 50%'] = df['Trump Win Probability (%)'] - 50

# Centered bar chart
fig, ax = plt.subplots(figsize=(10, 6))
colors = ['red' if x > 0 else 'blue' for x in df['Distance from 50%']]
ax.barh(df['State'], df['Distance from 50%'], color=colors)
ax.axvline(0, color='black')
ax.set_xlabel('Distance from 50% (Trump Winning Probability)')
ax.set_title('Probability of Trump Win, Centered from 50%')
plt.tight_layout()
plt.show()

Kamala is predicted to win he remaining of th states and is givn fairly high chances of locking these up. despite only winning Nevada, North Carolina and Pennsilvania by under a percntag point in the model Kamalas probablity of winning thos eelectoral votes is vry high. This is due to the small standard dviations shown in table x making he prdicted rul of the states leaning democrat being mor likely.
Wisconson is shown to be her srongest and most likley victory of the swing states. 
Overall this givs Harris 51 of the 93 swing state votes his results in a win for the democrats and the first female US president!

## Trump overall probability
-- Taking he probabilities from the model above we can look at the outcomes of the 207 possible combinations of wins and loses for Trump in these swing states adn deduce his overall probability of reaching 270 electoral votes. Basd on the modles probabilities his current expected votes is 41.05, below the threshold of 51 he needs based on our assumptions.

Seperating the outcomes into those that give trump the required electoral college votes, we can obtain  probability that trump ceizes a second term.

Using a probaboiility theory, and assuming independence among the states, in our approach. When accounting for all possible state outcome combinations, the model calculates that Trump has approximately a 22.5% chance of securing at least 51 electoral votes from these key swing states. This probability aggregates the likelihoods across all combinations where the sum of electoral votes meets or exceeds the 51-vote threshold. The analysis reveals that despite high individual probabilities in certain states, the overall chance of Trump amassing sufficient electoral votes from the swing states is less than one in four.

@@

# Apendix
(include some detail on how we are getting rid of biass. and focus on how we are spcfically uing stratifid and clustr sampling. how we will reach ou. how we will gt rsponce.)

@@

## what is included in the poll?

## Methodology and why?
